---
title: "Using predictive margins"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using predictive margins}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Libraries

```{r setup}
library(tidyverse)
library(ranger)
library(data.table)
library(predictiveMargins)
```


# Data

use the `written_genitives` dataset from the paper. We'll convert the relevant columns to factors.

```{r}
written_genitives <- written_genitives |>
  mutate(
    Type = as.factor(Type),
    Response = as.numeric(Type) - 1,
    Genre = as.factor(Genre),
    Possessor_Animacy3 = as.factor(Possessor_Animacy3),
    Final_Sibilant = as.factor(Final_Sibilant),
    Possessor_NP_Type = as.factor(Possessor_NP_Type)
  )
glimpse(written_genitives)
```
# Fit random forest model

Define the formula.

```{r}
fmla <- Type ~ Genre + Possessor_Animacy3 + Possessor_NP_Type +
  Final_Sibilant + Possessor_Length + Possessor_Thematicity
```

Fit the model. Here we use a probability forest rather than a classification forest by setting `probability = TRUE`. In a real case we would tune the model's hyperparameters, e.g. with the `{tidymodels}` or `{tuneRanger}` packages, but this does not affect the functions in `{predictiveMargins}`.

```{r}
rf1 <- ranger(
  fmla,
  data = written_genitives,
  num.trees = 1000,
  mtry = 3,
  probability = TRUE,
  respect.unordered.factors = "partition"
)
```

# Get marginal predictions

To get the marginal predictions of the trees, we use the `marginal_predictions()` function. The two obligatory arguments are the model `rf1` and the dataset used to train the model.

```{r}
marginal_preds <- marginal_predictions(rf1, written_genitives)
```
Additional optional arguments include:

- `num.trees`: the number of trees from which to take predictions. If this number is smaller than the total number of trees in the forest, a random sample is taken. The default is 500 trees. This is to keep the resulting datasets to a reasonable size, as forests may contain many thousands of trees.
- `n.breaks`: the number of breakpoints for binning continuous predictor variables. Default is 10. *Note that increasing this number can greatly impact speed and memory cost, especially when there are many other predictor variables.*
- `verbose`: If `TRUE` additional information will be printed about the dataset.

See the output of `marginal_preds`.

```{r}
names(marginal_preds)
```

## Average marginal predictions for categorical predictor

First we calculate the weighted average marginal predictions for the association of `Genre` with genitive construction. 

The default is the isolated weighting (`"iso"`), where the weights are based on the expected distributions of the feature combinations calculated from the marginal distributions of each peripheral predictor.

```{r}
genre_mar_avg_iso <- marginal_avg(marginal_preds, "Genre")
```
```{r}
genre_mar_avg_iso
```

now try the joint weighting.

```{r}
genre_mar_avg_joint <- marginal_avg(marginal_preds, "Genre", wt = "joint")
```
Now we can plot these calculations.

```{r}
bind_rows(
  genre_mar_avg_iso |>
    mutate(weighting = "isolated"),
  genre_mar_avg_joint |>
    mutate(weighting = "joint")
) |>
  ggplot(aes(x = Genre, y = mean_prob_s, color = weighting)) +
  geom_line(aes(group = weighting), position = position_dodge(width = .2)) +
  geom_pointrange(aes(ymin = lower, ymax = upper),
                  position = position_dodge(width = .2)) +
  theme_classic()
```

## Average marginal predictions for continuous predictor

```{r}
marginal_avg(marginal_preds, "Possessor_Thematicity", wt = "joint", verbose = F) |>
  ggplot(aes(Possessor_Thematicity, mean_prob_s)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = .3) +
  geom_line(aes(group = 1)) +
  theme_classic()
```

## Contrasts

```{r}
np_type_contrasts <- marginal_preds |>
  marginal_contrasts("Possessor_NP_Type")
```
```{r}
glimpse(np_type_contrasts)
```



```{r}
ggplot(np_type_contrasts, aes(x = mean_Possessor_NP_Type_contrast)) +
  geom_vline(xintercept = 0, color = "blue") +
  geom_histogram(color = "black", fill = "gray", bins = 50) +
  theme_classic()
```
```{r}
animacy_contrasts <- marginal_preds |>
  marginal_contrasts("Possessor_Animacy3")
```

```{r}
glimpse(animacy_contrasts)
```
```{r}
animacy_contrasts |> 
  summarise(across(6:8, ~ median(.x)))
```



```{r fig.width=6}
library(ggridges)

animacy_contrasts |> 
  pivot_longer(6:8, names_to = "contrast") |> 
  ggplot(aes(value, contrast)) +
  geom_density_ridges() +
  # stat_summary(geom = "pointrange", fun.data = "median_hilow") +
  theme_classic()
```

```{r}
genre_contrasts <- marginal_preds |>
  marginal_contrasts("Genre")
```

```{r}
glimpse(genre_contrasts)
```

We can plot these and color by the comparisons with Press (note the marginal averages above showed a distinct difference between Press and the rest).

```{r fig.width=6}
genre_contrasts |> 
  pivot_longer(6:15, names_to = "contrast") |> 
  mutate(
    color = if_else(grepl("Press", contrast), "press", "nonpress")
  ) |> 
  ggplot(aes(value, contrast, fill = color)) +
  geom_density_ridges() +
  # stat_summary(geom = "pointrange", fun.data = "median_hilow") +
  theme_classic()
```

All the comparisons with Press show long left tails. Because these reflect the difference in predicted probability of GENRE - Press, the left tails tell us that there are many combinations in which the *s*-genitive is predicted to be **less** likely in the stated genre than in Press texts.  










